---
title: "Project proposal"
author: "Team 8"
output: pdf_document
---

```{r load-packages, message = FALSE}
library(knitr)
library(tidyverse)
library(broom)
library(htmltools)
```

```{r setup, include = FALSE}
opts_chunk$set(echo=FALSE) # hide source code in the document
```

## 1. Introduction

## 2. Data

We are using five datasets to explore how the COVID-19 pandemic affected domestic violence trends across 25 countries. Our main dataset contains monthly reports of domestic violence. The other four datasets include lockdown policies, COVID-related deaths, unemployment rates, and public mental health indicators. 
Some of our datasets of different measurments, we are looking to see trends in every country as its own so the measurment does not matter nor needs to be normalized.
Also some of the datasets do not contain all the countries from our main dataset, but because we are working on 25 different countries and look for trends generally and not specify by country then it is acceptable to research some factors on a partly dataset of countries as long as we have more than 10-15 countries and their other factors exist and have a difference between them.

A full description of each dataset and the features we are using is provided in the README.md file in the `/data` folder and appears in the appendix of this document.

## 3. Preliminary results

## 4. Data analysis plan

## Appendix

### Data README

```{r include_data_readme, comment=''}
cat(readLines('../data/README.md'), sep = '\n')
```

### Source code

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
