
# load modeling packages
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lmtest)     # robust standard errors
library(sandwich)   # more robust standard errors  
library(plm)        # panel data models
library(stargazer)  # nice model output tables
library(car)        # diagnostic tests
library(broom)      # tidy model results
library(performance) # model evaluation

# print current working directory
cat("working in:", getwd(), "\n")
```

# load our prepared data
```{r data-load}
# simple loading - processed files should be in current directory
# this matches how you work in your other scripts

tryCatch({
  dv_enhanced <- read_csv("data/dv_analysis_enhanced.csv")
  country_profiles <- read_csv("data/country_typology.csv")
  message("loaded processed data from current directory")
}, error = function(e) {
  cat("could not find processed data files!\n")
  cat("make sure you've run the data preparation script first (01_enhanced_data_prep.Rmd)\n") 
  cat("looking for: dv_analysis_enhanced.csv and country_typology.csv\n")
  cat("current directory:", getwd(), "\n")
  stop("processed data files not found")
})

cat("loaded", nrow(dv_enhanced), "observations across", n_distinct(dv_enhanced$Entity), "countries\n")
```

# enhanced data preparation for modeling
```{r data-prep}
# preparing data specifically for our statistical models
# first let's check what variables we actually have
cat("columns in dv_enhanced:\n")
print(colnames(dv_enhanced))

# check if we have population data
if("population" %in% colnames(dv_enhanced)) {
  cat("found population column\n")
} else {
  cat("no population column found - will create dummy variable\n")
}

dv_model_ready <- dv_enhanced %>%
  mutate(
    Month_date = as.Date(Month_date),
    
    # ensure our key binary classification is clean
    policy_approach = case_when(
      policy_response_type %in% c("Economic_Focus", "Quick_Strong") ~ "Restrictive",
      policy_response_type %in% c("Gradual_Moderate", "Minimal_Response") ~ "Moderate",
      TRUE ~ NA_character_
    ),
    policy_approach = factor(policy_approach, levels = c("Moderate", "Restrictive")),
    
    # standardize continuous predictors for interpretability  
    lockdown_intensity_std = as.numeric(scale(lockdown_intensity_per_day)),
    economic_support_std = as.numeric(scale(Monthly_income_support)),
    death_rate_std = as.numeric(scale(Monthly_death_rates_per_million_people)),
    
    # create population proxy (since Entity is country name, not population)
    # use a simple country size proxy or set to 0 for simplicity
    log_population_scaled = 0,  # simplified - can remove from models if needed
    
    # time and lockdown variables
    lockdown_weeks = Total_lockdown_length / 7,
    lockdown_above_3weeks = as.numeric(lockdown_weeks > 3),
    high_death_rate = as.numeric(Monthly_death_rates_per_million_people > 50),
    
    # time trend
    time_trend = as.numeric(Month_date - min(Month_date, na.rm = TRUE)) / 30,
    
    # baseline measures - fix the baseline calculation
    baseline_dv_std = as.numeric(scale(replace_na(baseline_dv, mean(baseline_dv, na.rm = TRUE))))
  ) %>%
  filter(is_pandemic == 1, !is.na(policy_approach), !is.na(dv_change_from_baseline))

# check for missing values in key variables
cat("checking missing values in key variables:\n")
key_vars <- c("dv_change_from_baseline", "policy_approach", "lockdown_intensity_std", 
              "time_trend", "baseline_dv_std")
for(var in key_vars) {
  if(var %in% colnames(dv_model_ready)) {
    n_missing <- sum(is.na(dv_model_ready[[var]]))
    cat(var, ":", n_missing, "missing values\n")
  } else {
    cat(var, ": variable not found\n")
  }
}

cat("model-ready dataset:", nrow(dv_model_ready), "pandemic observations\n")
cat("complete cases for basic model:", sum(complete.cases(dv_model_ready[key_vars])), "\n")
```

# balance check between policy groups  
```{r balance-check}
# important to check if our restrictive vs moderate groups are comparable
balance_table <- dv_model_ready %>%
  group_by(policy_approach) %>%
  summarise(
    n_obs = n(),
    n_countries = n_distinct(Entity),
    
    # dv outcomes
    mean_dv_change = mean(dv_change_from_baseline, na.rm = TRUE),
    sd_dv_change = sd(dv_change_from_baseline, na.rm = TRUE),
    
    # policy characteristics  
    mean_lockdown_intensity = mean(lockdown_intensity_per_day, na.rm = TRUE),
    mean_lockdown_weeks = mean(lockdown_weeks, na.rm = TRUE),
    pct_above_3weeks = mean(lockdown_above_3weeks, na.rm = TRUE) * 100,
    mean_economic_support = mean(Monthly_income_support, na.rm = TRUE),
    
    # health impacts
    mean_death_rate = mean(Monthly_death_rates_per_million_people, na.rm = TRUE),
    .groups = "drop"
  )

print("balance between policy approach groups:")
print(balance_table)

# this shows us that our classification captures meaningful differences
```

# setup panel data structure
```{r panel-setup}
# creating panel data frame for country fixed effects models
panel_data <- pdata.frame(dv_model_ready,
                         index = c("Entity", "Month_date"), 
                         drop.index = FALSE)

cat("panel data ready with", length(unique(index(panel_data, "id"))), "countries\n")
```

# systematic model building progression
```{r ols-models}
# building models step by step to understand what matters most
# simplified to avoid missing value issues

# model 1: basic controls only
model1_baseline <- lm(dv_change_from_baseline ~ 
                       baseline_dv_std +
                       time_trend,
                     data = dv_model_ready)

# model 2: add our key policy classification
model2_policy <- lm(dv_change_from_baseline ~
                     baseline_dv_std +
                     time_trend +
                     policy_approach,
                   data = dv_model_ready)

# model 3: add policy intensity measures
model3_intensity <- lm(dv_change_from_baseline ~
                        baseline_dv_std +
                        time_trend +
                        policy_approach +
                        lockdown_intensity_std,
                      data = dv_model_ready)

# model 4: add threshold effects (our key finding)
model4_thresholds <- lm(dv_change_from_baseline ~
                         baseline_dv_std +
                         time_trend +
                         policy_approach +
                         lockdown_intensity_std +
                         lockdown_above_3weeks,
                       data = dv_model_ready)

# model 5: add economic support
model5_economic <- lm(dv_change_from_baseline ~
                       baseline_dv_std +
                       time_trend + 
                       policy_approach +
                       lockdown_intensity_std +
                       lockdown_above_3weeks +
                       economic_support_std,
                     data = dv_model_ready)

# model 6: our final specification with interaction
model6_final <- lm(dv_change_from_baseline ~
                    baseline_dv_std +
                    time_trend +
                    policy_approach +
                    lockdown_intensity_std +
                    lockdown_above_3weeks +
                    policy_approach:lockdown_intensity_std,
                  data = dv_model_ready)

cat("built 6 progressive models\n")

# check if models worked
for(i in 1:6) {
  model_name <- paste0("model", i, if(i==1) "_baseline" else if(i==2) "_policy" else if(i==3) "_intensity" else if(i==4) "_thresholds" else if(i==5) "_economic" else "_final")
  model_obj <- get(model_name)
  cat(model_name, "- observations used:", nobs(model_obj), "\n")
}
```

# panel data models for robustness
```{r panel-models}
# these models control for unobserved country characteristics

# fixed effects model (removes country-specific factors)
model_fe <- plm(dv_change_from_baseline ~
                 time_trend +
                 policy_approach +
                 lockdown_intensity_std +
                 lockdown_above_3weeks,
               data = panel_data,
               model = "within")

# random effects model
model_re <- plm(dv_change_from_baseline ~
                 baseline_dv_std +
                 time_trend +
                 policy_approach +
                 lockdown_intensity_std +
                 lockdown_above_3weeks,
               data = panel_data,
               model = "random")

# hausman test to choose between fe and re
tryCatch({
  hausman_test <- phtest(model_fe, model_re)
  print("hausman test (fe vs re):")
  print(hausman_test)
}, error = function(e) {
  cat("could not run hausman test:", e$message, "\n")
})
```

# robust standard errors for all models
```{r robust-se}
# clustering standard errors by country to account for within-country correlation

# function to get clustered standard errors
get_robust_se <- function(model) {
  if (inherits(model, "lm")) {
    vcovCL(model, cluster = ~Entity, type = "HC1")
  } else {
    vcovHC(model, type = "HC1", cluster = "group") 
  }
}

# create list of our key models
models_list <- list(
  "Baseline" = model1_baseline,
  "Policy" = model2_policy, 
  "Intensity" = model3_intensity,
  "Thresholds" = model4_thresholds,
  "Economic" = model5_economic,
  "Final" = model6_final
)

# get robust standard errors for each (only for models that worked)
robust_results <- list()
for(name in names(models_list)) {
  tryCatch({
    robust_results[[name]] <- coeftest(models_list[[name]], vcov = get_robust_se(models_list[[name]]))
    cat("robust se calculated for", name, "\n")
  }, error = function(e) {
    cat("error with robust se for", name, ":", e$message, "\n")
  })
}

cat("calculated robust standard errors for", length(robust_results), "models\n")
```

# model diagnostics and comparison
```{r diagnostics}
# checking our models meet key assumptions

# variance inflation factors for multicollinearity (only if final model worked)
tryCatch({
  vif_final <- vif(model6_final)
  print("vif values (final model):")
  print(vif_final[vif_final > 2])  # flag high vif
}, error = function(e) {
  cat("could not calculate vif:", e$message, "\n")
})

# model comparison table (only for models that worked)
model_comparison <- tibble(
  Model = character(0),
  R2 = numeric(0),
  Adj_R2 = numeric(0),
  AIC = numeric(0),
  BIC = numeric(0),
  RMSE = numeric(0),
  N_obs = numeric(0)
)

for(name in names(models_list)) {
  tryCatch({
    model <- models_list[[name]]
    model_comparison <- model_comparison %>%
      add_row(
        Model = name,
        R2 = summary(model)$r.squared,
        Adj_R2 = summary(model)$adj.r.squared,
        AIC = AIC(model),
        BIC = BIC(model),
        RMSE = sqrt(mean(residuals(model)^2)),
        N_obs = nobs(model)
      )
  }, error = function(e) {
    cat("could not get stats for model", name, ":", e$message, "\n")
  })
}

if(nrow(model_comparison) > 0) {
  model_comparison <- model_comparison %>%
    mutate(
      AIC_delta = AIC - min(AIC, na.rm = TRUE),
      BIC_delta = BIC - min(BIC, na.rm = TRUE)
    )
  
  print("model performance comparison:")
  print(model_comparison)
} else {
  cat("no models succeeded\n")
}
```

# threshold analysis - our key substantive finding
```{r threshold-analysis}
# examining the 3-week lockdown threshold effect by policy approach

threshold_effects <- dv_model_ready %>%
  mutate(lockdown_duration = ifelse(lockdown_above_3weeks == 1, "Above 3 weeks", "3 weeks or less")) %>%
  group_by(policy_approach, lockdown_duration) %>%
  summarise(
    n_obs = n(),
    mean_dv_change = mean(dv_change_from_baseline, na.rm = TRUE),
    se_dv_change = sd(dv_change_from_baseline, na.rm = TRUE) / sqrt(n()),
    median_lockdown_weeks = median(lockdown_weeks, na.rm = TRUE),
    
    # economic support context
    mean_economic_support = mean(Monthly_income_support, na.rm = TRUE),
    
    # health context  
    mean_death_rate = mean(Monthly_death_rates_per_million_people, na.rm = TRUE),
    .groups = "drop"
  )

print("threshold effects by policy approach:")
print(threshold_effects)

# calculate the key finding: difference in dv change above vs below 3 weeks
threshold_summary <- threshold_effects %>%
  select(policy_approach, lockdown_duration, mean_dv_change) %>%
  pivot_wider(names_from = lockdown_duration, values_from = mean_dv_change) %>%
  mutate(threshold_effect = `Above 3 weeks` - `3 weeks or less`)

print("threshold effect (above 3 weeks - below 3 weeks):")
print(threshold_summary)
```

# extract final model results for visualization
```{r final-results}
# getting clean results from our best model for plotting

# check which models worked and use the best one
if(exists("model6_final") && !is.null(model6_final)) {
  best_model <- model6_final
  model_name <- "Final Model"
} else if(exists("model4_thresholds") && !is.null(model4_thresholds)) {
  best_model <- model4_thresholds  
  model_name <- "Thresholds Model"
} else if(exists("model2_policy") && !is.null(model2_policy)) {
  best_model <- model2_policy
  model_name <- "Policy Model"
} else {
  stop("no models worked successfully")
}

cat("using", model_name, "for final results\n")

# get coefficients with confidence intervals
final_effects <- tidy(best_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    effect_direction = ifelse(estimate > 0, "Increase DV", "Decrease DV"),
    effect_size_cat = case_when(
      abs(estimate) < 5 ~ "Small",
      abs(estimate) < 10 ~ "Medium", 
      TRUE ~ "Large"
    )
  )

print("final model key results:")
print(final_effects %>% select(term, estimate, conf.low, conf.high, p.value, significant))

# interaction effects specifically
interaction_effects <- final_effects %>%
  filter(str_detect(term, ":"))

if(nrow(interaction_effects) > 0) {
  print("interaction effects:")
  print(interaction_effects)
} else {
  cat("no interaction effects in this model\n")
}
```

# save model results for visualization
```{r save-results}
# saving everything we need for the visualization script

# save individual model objects and results that actually worked
save_objects <- list(
  "final_effects" = if(exists("final_effects")) final_effects else NULL,
  "threshold_effects" = if(exists("threshold_effects")) threshold_effects else NULL,
  "model_comparison" = if(exists("model_comparison")) model_comparison else NULL,
  "dv_model_ready" = dv_model_ready,
  "balance_table" = if(exists("balance_table")) balance_table else NULL
)

# add the best model
if(exists("best_model")) {
  save_objects[["best_model"]] <- best_model
}

# save as RData file
save(list = names(save_objects)[!sapply(save_objects, is.null)], 
     file = "enhanced_statistical_models.RData")

# also save as csv files for easier access (only if they exist)
if(exists("final_effects") && !is.null(final_effects)) {
  write_csv(final_effects, "final_model_coefficients.csv")
}
if(exists("threshold_effects") && !is.null(threshold_effects)) {
  write_csv(threshold_effects, "threshold_analysis_results.csv")
}
if(exists("model_comparison") && !is.null(model_comparison)) {
  write_csv(model_comparison, "model_comparison_table.csv")
}

cat("model results saved successfully\n")
cat("next step: run the visualization notebook\n")
```

# quick summary of key findings
```{r summary}
cat("=== key findings summary ===\n")
cat("1. policy approach effect:\n")

policy_effect <- balance_table %>%
  select(policy_approach, mean_dv_change) %>%
  pivot_wider(names_from = policy_approach, values_from = mean_dv_change) %>%
  mutate(difference = Restrictive - Moderate)

cat("   restrictive:", round(policy_effect$Restrictive, 1), "% change\n")
cat("   moderate:", round(policy_effect$Moderate, 1), "% change\n") 
cat("   difference:", round(policy_effect$difference, 1), "percentage points\n\n")

cat("2. lockdown threshold effect (3 weeks):\n")
print(threshold_summary)

cat("\n3. model performance:\n")
cat("   final model adj r²:", round(model_comparison$Adj_R2[model_comparison$Model == "Final"], 3), "\n")
cat("   improvement from baseline:", 
    round(model_comparison$Adj_R2[model_comparison$Model == "Final"] - 
          model_comparison$Adj_R2[model_comparison$Model == "Baseline"], 3), "\n")
```

# threshold analysis - our key substantive finding
```{r threshold-analysis}
# examining the 3-week lockdown threshold effect by policy approach

threshold_effects <- dv_model_ready %>%
  mutate(lockdown_duration = ifelse(lockdown_above_3weeks == 1, "Above 3 weeks", "3 weeks or less")) %>%
  group_by(policy_approach, lockdown_duration) %>%
  summarise(
    n_obs = n(),
    mean_dv_change = mean(dv_change_from_baseline, na.rm = TRUE),
    se_dv_change = sd(dv_change_from_baseline, na.rm = TRUE) / sqrt(n()),
    median_lockdown_weeks = median(lockdown_weeks, na.rm = TRUE),
    
    # economic support context
    mean_economic_support = mean(Monthly_income_support, na.rm = TRUE),
    
    # health context  
    mean_death_rate = mean(Monthly_death_rates_per_million_people, na.rm = TRUE),
    .groups = "drop"
  )

print("threshold effects by policy approach:")
print(threshold_effects)

# calculate the key finding: difference in dv change above vs below 3 weeks
threshold_summary <- threshold_effects %>%
  select(policy_approach, lockdown_duration, mean_dv_change) %>%
  pivot_wider(names_from = lockdown_duration, values_from = mean_dv_change) %>%
  mutate(threshold_effect = `Above 3 weeks` - `3 weeks or less`)

print("threshold effect (above 3 weeks - below 3 weeks):")
print(threshold_summary)

---