---
title: "Project proposal"
author: "Team 8"
output: pdf_document
---

```{r load-packages, message = FALSE}
library(knitr)
library(tidyverse)
library(broom)
library(htmltools)
```

```{r setup, include = FALSE}
opts_chunk$set(echo=FALSE) # hide source code in the document
```

## 1. Introduction

## 2. Data

We are using four datasets to explore how the COVID-19 pandemic affected domestic violence trends across 25 countries. Our main dataset contains monthly reports of domestic violence. The other three datasets include lockdown policies, COVID-related deaths and unemployment rates. 
Some of our datasets of different measurments, we are looking to see trends in every country as its own so the measurment does not matter nor needs to be normalized.
Also some of the datasets do not contain all the countries from our main dataset, but because we are working on 25 different countries and look for trends generally and not specify by country then it is acceptable to research some factors on a partly dataset of countries as long as we have more than 10-15 countries and their other factors exist and have a difference between them.

A full description of each dataset and the features we are using is provided in the README.md file in the `/data` folder and appears in the appendix of this document.

## 3. Preliminary results
In this analysis, we examine the relationship between trends in deaths per 100,000 people and reported violence incidents during the first months of the COVID-19 pandemic.

We plotted two time series graphs side-by-side:
One shows the monthly deaths per 100,000 people
The other shows the average violence indicator over the same period

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)


df <- read_csv("merged_deaths_violence_final.csv")


df_summary <- df %>%
  mutate(year_month = as.Date(paste0(year_month, "-01"))) %>%
  group_by(year_month) %>%
  summarise(
    avg_deaths = mean(monthly_avg_deaths, na.rm = TRUE),
    avg_violence = mean(violence_indicator, na.rm = TRUE)
  ) %>%
  mutate(deaths_per_100k = avg_deaths * 100000)


p1 <- ggplot(df_summary, aes(x = year_month, y = deaths_per_100k, group = 1)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "firebrick", size = 2) +
  labs(title = "Deaths per 100,000 Over Time",
       x = "Month", y = "Deaths per 100K") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(df_summary, aes(x = year_month, y = avg_violence, group = 1)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Violence Indicator Over Time",
       x = "Month", y = "Violence Indicator") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


(p1 + p2) +
  plot_annotation(title = "Comparison Between Deaths and Violence Trends")



```
The graphs reveal that during the COVID-19 outbreak peak around April 2020, the number of deaths per 100,000 increased sharply. In contrast, the violence indicator shows a sharp decrease at the exact same time. After April, we observe an inverse pattern: while deaths begin to decline, violence starts to rise again.

These opposite trends suggest a potential negative association between mortality levels and violence during the pandemic — possibly reflecting factors such as lockdown severity, reduced social interaction, or underreporting of violence



This analysis examines the global trends in unemployment per 1000 people and violence incidents during the early phase of the COVID-19 pandemic

We present two side-by-side time series plots
The left plot shows the average global unemployment rate per 1000 people
The right plot shows the average global violence indicator over time

```{r}
library(dplyr)
library(readr)
library(stringr)


merge <- read_csv("merged_deaths_violence_final.csv", show_col_types = FALSE) %>%
  mutate(Entity = str_to_lower(Entity),
         year_month = as.character(year_month))

unemp <- read_csv("UNE_TUNE_SEX_AGE_NB_M-filtered-2025-05-18.csv", show_col_types = FALSE) %>%
  mutate(
    Entity = str_to_lower(`ref_area.label`),
    year_month = str_replace(time, "M", "-")  # הפיכת 2020M12 ל־ 2020-12
  )

unemp_summary <- unemp %>%
  group_by(Entity, year_month) %>%
  summarise(unemployment_rate = mean(obs_value, na.rm = TRUE), .groups = "drop")


merged_final <- merge %>%
  left_join(unemp_summary, by = c("Entity", "year_month"))

write_csv(merged_final, "merged_with_unemployment.csv")

```

```{r}

library(dplyr)
library(readr)
library(ggplot2)
library(gridExtra)

df <- read_csv("merged_with_unemployment.csv")

df_filtered <- df %>% filter(!is.na(unemployment_rate))


monthly_avg <- df_filtered %>%
  group_by(year_month) %>%
  summarise(
    avg_unemployment = mean(unemployment_rate, na.rm = TRUE),
    avg_violence = mean(violence_indicator, na.rm = TRUE),
    .groups = "drop"
  )

monthly_avg$year_month <- as.Date(paste0(monthly_avg$year_month, "-01"))


p1 <- ggplot(monthly_avg, aes(x = year_month, y = avg_unemployment)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "orange", size = 2) +
  labs(title = "Average Unemployment Rate (with data)",
       x = "Month", y = "Unemployment per 1,000 people") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(monthly_avg, aes(x = year_month, y = avg_violence)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Average Violence Indicator (with unemployment data)",
       x = "Month", y = "Violence Indicator") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


grid.arrange(p1, p2, ncol = 2, top = "Trends in Unemployment and Violence (with overlapping data only)")


```
Although the relationship between the two variables is less clearly defined compared to previous analyses we still observe a notable shift around April 2020 when the COVID-19 pandemic began to peak globally
During this time the unemployment rate begins a steady increase that continues in the following months
In contrast the violence indicator reaches its lowest point in April followed by a gradual upward trend

This simultaneous change suggests that the early stages of the pandemic may have impacted both economic and social stability
While the trends are not perfectly inversely correlated the timing of their shifts hints at a possible indirect connection between rising unemployment and increases in violence




## 4. Data analysis plan

Response (Y) Variable:

Violence indicator: Our primary outcome measure representing reported domestic violence incidents

Explanatory (X) Variables:

-COVID mortality rates: Monthly deaths per 100,000 people
-Lockdown severity: Categorized levels of government restrictions
-Unemployment rates: Monthly unemployment figures per 1,000 people

Comparison Groups
Our analysis will leverage several comparison groups to identify patterns and differences:
1. Geographic Comparison Groups

-Continental regions: Group countries by continent (Europe, Asia, Americas, etc.)
-Economic development level: High-income vs. middle-income vs. low-income countries
-Cultural regions: Group countries with similar cultural characteristics

2. Policy Response Comparison Groups

Lockdown severity: Group countries by strictness of lockdown policies

-High restriction (full lockdowns, stay-at-home orders)
-Medium restriction (partial closures, curfews)
-Low restriction (advisory measures, minimal restrictions)


Policy timing: Early responders vs. delayed responders
Policy duration: Short-term vs. long-term restrictions

3. Temporal Comparison Groups

Pandemic phases:

-Pre-pandemic (before March 2020)
-First wave (March-June 2020)
-Intermediate period (July-October 2020)
-Second wave (November 2020 onwards)

Seasonal periods: To control for seasonal variations in violence reporting

4. Impact-Based Comparison Groups

-Mortality impact: Countries with high vs. medium vs. low COVID-19 death rates
-Economic impact: Countries with high vs. medium vs. low unemployment increases
-Combined impact: Create a composite score of health and economic impacts

Analytical Methods

Expected Findings from Group Comparisons
We anticipate discovering:

Regional variations: European countries may show different patterns than Asian or American countries due to cultural factors and social support systems
Policy impact differences: High-restriction countries likely showing more pronounced V-shaped violence trends (initial decrease followed by increase)
Temporal differences: First wave vs. second wave effects may differ as countries adapted response strategies
Economic moderators: High unemployment countries may show stronger relationships between lockdowns and violence

Teamwork Division
Our analysis work will be divided as follows:

Data Preparation and Feature Engineering:

Team member 1: Clean and merge datasets, handle missing values
Team member 2: Create derived features, normalize variables


Comparison Group Analysis:

Team member 3: Conduct ANOVA, t-tests, and statistical comparisons
Team member 4: Design and implement visualization strategies for group comparisons


Statistical Modeling:

Team member 1: Implement region-specific models
Team member 2: Develop policy-group models
Team member 3: Create and evaluate temporal models
Team member 4: Coordinate model evaluation and comparison across groups


Results and Report Preparation:

All team members will collaborate on interpreting results and preparing the final report
I will lead the synthesis of findings across groups and the development of recommendations



## Appendix

### Data README

```{r include_data_readme, comment=''}
cat(readLines('../data/README.md'), sep = '\n')
```

### Source code

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
