---
title: "Project proposal"
author: "Team 8"
output: pdf_document
---

```{r load-packages, message = FALSE}
library(knitr)
library(tidyverse)
library(broom)
library(htmltools)
```

```{r setup, include = FALSE}
opts_chunk$set(echo=FALSE) # hide source code in the document
```

## 1. Introduction

## 2. Data

We are using four datasets to explore how the COVID-19 pandemic affected domestic violence trends across 25 countries. Our main dataset contains monthly reports of domestic violence. The other three datasets include lockdown policies, COVID-related deaths and unemployment rates. 
Some of our datasets of different measurments, we are looking to see trends in every country as its own so the measurment does not matter nor needs to be normalized.
Also some of the datasets do not contain all the countries from our main dataset, but because we are working on 25 different countries and look for trends generally and not specify by country then it is acceptable to research some factors on a partly dataset of countries as long as we have more than 10-15 countries and their other factors exist and have a difference between them.

A full description of each dataset and the features we are using is provided in the README.md file in the `/data` folder and appears in the appendix of this document.

## 3. Preliminary results
In this analysis, we examine the relationship between trends in deaths per 100,000 people and reported violence incidents during the first months of the COVID-19 pandemic.

We plotted two time series graphs side-by-side:
One shows the monthly deaths per 100,000 people
The other shows the average violence indicator over the same period

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)


df <- read_csv("merged_deaths_violence_final.csv")


df_summary <- df %>%
  mutate(year_month = as.Date(paste0(year_month, "-01"))) %>%
  group_by(year_month) %>%
  summarise(
    avg_deaths = mean(monthly_avg_deaths, na.rm = TRUE),
    avg_violence = mean(violence_indicator, na.rm = TRUE)
  ) %>%
  mutate(deaths_per_100k = avg_deaths * 100000)


p1 <- ggplot(df_summary, aes(x = year_month, y = deaths_per_100k, group = 1)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "firebrick", size = 2) +
  labs(title = "Deaths per 100,000 Over Time",
       x = "Month", y = "Deaths per 100K") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(df_summary, aes(x = year_month, y = avg_violence, group = 1)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Violence Indicator Over Time",
       x = "Month", y = "Violence Indicator") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


(p1 + p2) +
  plot_annotation(title = "Comparison Between Deaths and Violence Trends")



```
The graphs reveal that during the COVID-19 outbreak peak around April 2020, the number of deaths per 100,000 increased sharply. In contrast, the violence indicator shows a sharp decrease at the exact same time. After April, we observe an inverse pattern: while deaths begin to decline, violence starts to rise again.

These opposite trends suggest a potential negative association between mortality levels and violence during the pandemic — possibly reflecting factors such as lockdown severity, reduced social interaction, or underreporting of violence



This analysis examines the global trends in unemployment per 1000 people and violence incidents during the early phase of the COVID-19 pandemic

We present two side-by-side time series plots
The left plot shows the average global unemployment rate per 1000 people
The right plot shows the average global violence indicator over time

```{r}
library(dplyr)
library(readr)
library(stringr)


merge <- read_csv("merged_deaths_violence_final.csv", show_col_types = FALSE) %>%
  mutate(Entity = str_to_lower(Entity),
         year_month = as.character(year_month))

unemp <- read_csv("UNE_TUNE_SEX_AGE_NB_M-filtered-2025-05-18.csv", show_col_types = FALSE) %>%
  mutate(
    Entity = str_to_lower(`ref_area.label`),
    year_month = str_replace(time, "M", "-")  # הפיכת 2020M12 ל־ 2020-12
  )

unemp_summary <- unemp %>%
  group_by(Entity, year_month) %>%
  summarise(unemployment_rate = mean(obs_value, na.rm = TRUE), .groups = "drop")


merged_final <- merge %>%
  left_join(unemp_summary, by = c("Entity", "year_month"))

# שמירה לקובץ חדש
write_csv(merged_final, "merged_with_unemployment.csv")

```

```{r}

monthly_avg$year_month <- as.Date(paste0(monthly_avg$year_month, "-01"))


monthly_avg$month_label <- format(monthly_avg$year_month, "%b")
monthly_avg$month_label <- factor(monthly_avg$month_label,
                                  levels = format(sort(unique(monthly_avg$year_month)), "%b"))


p1 <- ggplot(monthly_avg, aes(x = month_label, y = avg_unemployment, group = 1)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "orange", size = 2) +
  labs(title = "Unemployment Over Time",
       x = "Month", y = "Unemployment per 1,000 people") +
  theme_minimal()


p2 <- ggplot(monthly_avg, aes(x = month_label, y = avg_violence, group = 1)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Violence Indicator Over Time",
       x = "Month", y = "Violence Indicator") +
  theme_minimal()


grid.arrange(p1, p2, ncol = 2, top = "Trends in Unemployment and Violence During the Pandemic")


```
Although the relationship between the two variables is less clearly defined compared to previous analyses we still observe a notable shift around April 2020 when the COVID-19 pandemic began to peak globally
During this time the unemployment rate begins a steady increase that continues in the following months
In contrast the violence indicator reaches its lowest point in April followed by a gradual upward trend

This simultaneous change suggests that the early stages of the pandemic may have impacted both economic and social stability
While the trends are not perfectly inversely correlated the timing of their shifts hints at a possible indirect connection between rising unemployment and increases in violence

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)

# === Load and prepare data ===
dv <- read_csv("merged_with_unemployment.csv") %>%
  rename(country = Entity) %>%
  mutate(month = as.Date(paste0(year_month, "-01"))) %>%
  filter(Indicator == "Sexual violence or physical assault by IPFM*  (domestic violence): Total number of offences") %>%
  mutate(country = toupper(country))

lockdowns <- read_csv("greece_spain_newzealand_lockdowns.csv") %>%
  mutate(country = toupper(country),
         month = as.Date(paste0(year_month, "-01")))

selected_countries <- c("GREECE", "SPAIN", "NEW ZEALAND")

# Merge & clean
merged <- dv %>%
  filter(country %in% selected_countries) %>%
  left_join(lockdowns, by = c("country", "month")) %>%
  mutate(lockdown_length = replace_na(lockdown_length, 0)) %>%
  filter(format(month, "%Y-%m") != "2020-08")


# Clean and structure data
library(RColorBrewer)

# Assuming 'merged' is your combined dataset
plot_df <- merged %>%
  filter(format(month, "%Y-%m") != "2020-08") %>%
  mutate(
    country = factor(country, levels = c("GREECE", "SPAIN", "NEW ZEALAND")),
    month = as.Date(month)
  ) %>%
  group_by(country) %>%
  arrange(month) %>%
  mutate(
    dv_index = violence_indicator / first(violence_indicator) * 100
  ) %>%
  ungroup()

# Define a colorblind-friendly palette
cb_palette <- brewer.pal(n = 3, name = "Set2")

# Create the line plot
ggplot(plot_df, aes(x = month, y = dv_index, color = country)) +
  geom_line(size = 1) +
  geom_point(aes(size = lockdown_length), alpha = 0.8) +
  scale_color_manual(values = cb_palette) +
  scale_size_continuous(range = c(2, 6), name = "Lockdown Days") +
  scale_x_date(date_labels = "%b %Y", breaks = "1 month") +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Monthly Domestic Violence Trends During COVID-19",
    subtitle = "Lines represent DV offenses; point size indicates lockdown duration",
    x = "Month",
    y = "DV Index (First Month = 100)",
    color = "Country"
  ) +
  theme_minimal(base_size = 7) +
  theme(
    # Legend styling (already working)
    legend.position = "top",
    legend.box.margin = margin(t = 10, r = 5, b = 10, l = 0),
    legend.margin = margin(t = 3, b = 3),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 8),
    
    # Axes text (tick labels)
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
  
    # Axes titles
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold"),
  
    # Titles
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5)
  )






```
In this graph we compare three countries that handled COVID-19 in a different way through national lockdowns, Spain which had serious and long lockdowns, Greece which had partial and short lockdowns and New Zealand which had no lockdowns during this time, although it is only three countries we can see the domestic violence rates trends are very different and we would like to continue our research along with the other factors to fully understand what might cause the rise or fall of domestic violence during the beginning of the pandemic.



## 4. Data analysis plan



## Appendix

### Data README

```{r include_data_readme, comment=''}
cat(readLines('../data/README.md'), sep = '\n')
```

### Source code

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
